Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory

### 2018-02-27 Cars/paper.tex
Paper Revision

@@

\title{Online Updating of Variational Bayes for Forecasts of Vehicle Trajectory with Heterogeneous Driver Behaviour}
\begin{document}
\maketitle
\section{Introduction}

@@

\item Self driving Cars are becoming more common, eg Waymo, Tesla, GM.
\item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the Self Driving Car.
\item While Self Driving Cars reduce collisions due to human error, safety can be further improved by better forecasts of the trajectory of nearby human controlled vehicles.
\item These models use deep neural networks (DNN) to extract the properties of surrounding vehicles from sensors (eg. Cameras and LiDAR), such as their position, velocity, and steering angle (Woo et al. 2016a; Tian, Pei, Jana, and Ray 2017).
\item The Self Driving Car forecasts point estimates of the future trajectory of these surrounding vehicles based on the DNN results.
\item These can assume the surrounding vehicles will maintain their current angle and velocity (Gindele, Brechtel, Dillmann 2010; Waymo Safety Report 2017; Bautista 2017), but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines (Woo et al. 201b6; Ding et al. 2012; Woo et al. 2017; Geng et al. 2017).

@@

\itme An important part of this modelling is to capture the uncertainty of car trajectories, as the self-driving car must react to the possibility of collision even when that possibility is relatively small.

@@

\item This model implies that any new vehicle observed by the self driving car has its own parameter vector, related to previously observed vehicles through the shared prior.
\item The self driving car must be able to estimate the parameter vector of newly observed cars, and update this parameter vector as it observes additional data.

@@

\item Fit a hierarchical model with a large amount of vehicles to obtain the posterior distribution of each vehicles individual parameter vector and the shared distribution. This can occur before the self driving car is on the road.

@@

\item Self driving cars take in up to 1GB of data a second (several websites report this as quotes from different people eg. Tech CEO, but I don't know if that's good enough for an academic citaiton), so discarding old data is useful.

@@

\section{Data processing}

\begin{itemize}
\item Data is provided by the Next Generation Simulation (NGSIM) project, which observes cars for a 2000 feet section of the US 101 Highway.
\item Observations of x/y coordinates are provided every 100 ms (10 hz), with around 500-1000 observations per car.
\item Around 2000 of these cars changed lanes or stopped moving somewhere in the sample and are discarded.
\item There was some curvature in the road which means there are systematic movements in the car paths.
\item Modern cars are able to track lane marking and can calculate distance to other cars in relative co-ordinates (Thuy and Leon 2010), instead of relative to the Earth (ie. ignores curves)
\item Prior research fits a polynomial curve to detected lane markings to build a model of the road lane edges (Woo et. al. 2016) 
\item Instead, I fit splines to the positions of each car to build an estimate of the road midpoint. Relative co-ordinates are found separately for each lane using the movements of 100 cars / lane by the following procedure:
\begin{enumerate}
\item At each point in time $t$ calculate the total distance each car has travelled along the road using $d_{i, t} = \sum_{s=1}^t v_{i, s}$, where $v_{i, s}$ is the velocity of car $i$ at time $s$. 
\item Estimate independent smoothing splines of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
\item Use models to get $\hat{x}_{i, t}$ and $\hat{y}_{i, t}$ as estimates of the midpoint of the lane after travelling a distance $d_{i, t}$.
\item Define new coordinates 
\begin{align}
x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right) - \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
y^*_{i, t} &= d_{i, t} \label{yRel}
\end{align}
\end{enumerate}
\item (\ref{xRel}) \textit{really} needs a diagram.
\item These cars are not used for any further modelling, leaving 2873 cars travelling on a straightened out road with no noticable systematic curvature.
\end{itemize}

\section{Hierarchical Motion Model}

The position of any vehicle at time $t$ can be fully determined by its initial position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the vehicle's velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
\begin{align}
x^*_{t} &= x^*_{t-1} + v_{t} \cos(\delta_{t}) \label{xEq} \\
y^*_{t} &= y^*_{t-1} + v_{t} \sin(\delta_{t}) \label{yEq},
\end{align}
where $\delta_{t} = \pi/2$ denotes that the car has no lateral relative movement and $x^*_{t} = x^*_{t-1}$.
From this relationship, the driver's key inputs to motion can be extracted by
\begin{align}
\delta_{t} &= \tan^{-1}\left(\frac{(y^*_{t} - y^*_{t-1})}{(x^*_{t} - x^*_{t-1})} \right) \label{dEq} \\
v_{t} &= \sqrt{(x^*_{t} - x^*_{t-1})^2 + (y^*_{t} - y^*_{t-1})^2} \label{vEq} \\
a_{t} &= v_{t} - v_{t-1}. \label{aEq}
\end{align}
where $a_{t}$ denotes the acceleration at time $t$.

Auto-regressive processes of orders $p$ and $q$ for $a_t$ and  $\delta_t$ are used to model the level of persistance and variation of driver actions, with
\begin{align}
a_{t} &= \sum_{j = 1}^p \phi_{j} a_{t-j} + \sigma_{\epsilon} \epsilon_{t} \label{aAR} \\
\delta_{t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{t-j} - \pi/2) + \sigma_{\eta} \eta_{t} \label{dAR}
\end{align}
where $\epsilon_{t}$ and $\eta_t$ are independently and identically distributed according to a standard normal distribution. The unconditional means of $a_t$ and $\delta_t$ are set to zero and $\pi/2$ respectively, corresponding to forward motion at constant velocity. The parameter vector to be estimated is collected as 
\begin{equation*}
\label{thetaVec}
\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\} \in \mathbb{R}^{p + q + 2}.
\end{equation*}

Multiple groups of differing driver behaviour can be allowed while still retaining heterogeneity within each group by jointly estimating each vehicle driver's individual parameter vector, $\theta_i$, as part of a heirarchical model. In this case each $\theta_i$ is treated as an independent draw from a $K$ component mixture of multivariate normal distributions, augmenting each $\theta_i$ with a mixture component $k_i = 1, \dots, K$ such that 
\begin{equation}
\theta_i | k_i = j \sim N(\mu_j, \Sigma_j).
\end{equation}
For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with fixed mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with fixed degrees of freedom $\tau_j$ and scale matrix $\Psi_j$. $k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi} = \{\pi_{1}, \dots, \pi_{K}\}$, where $\boldsymbol{\pi}$ has a Dirichlet prior with fixed hyperparameters $\alpha = \alpha_j, j = 1 , \dots, K$. Defining $\beta = \{\mu_j, \Sigma_j, \pi_j, j = 1, \dots, K\}$, $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s} | s = 1, \dots, T\}$ and $z_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$ the posterior distribution $p(\theta_1, \dots, \theta_N, \beta | z_{1:N})$ can be estimated with MCMC methods.

\begin{figure}[ht]
\centering
\includegraphics[width = 0.95\textwidth]{HierSingleKDE.png}
\caption{Top: A Kernel Density Estimate applied to the individual posterior means for each $\theta_i | z_i, i = 1, \dots, 2000$ where $p = q = 2$, modelled independently without any hierarchical structure. Bottom: $p(\theta_i | \beta, z_{1:2000})$ for a hierarchical mixture model with $K = 6$ using the same vehicles, with each component of the mixture denoted by the coloured densities. The hierarchical model has captured the bimodality and skewness present in the independent posterior means.}
\label{fig:HierSingleKDE}
\end{figure}

\section{Variational Bayes}

\subsection{Background}

Variational Bayes (VB, Jordan et al. 1999) approximates a distribution $p(\theta | z)$ with a family $q(\theta)$, choosing the member of that family that minimises $KL(q(\theta) || p(\theta | z)$, where
\begin{equation}
\label{KL-div}
KL(q(\theta) || p(\theta | z) = \int_{\theta} q(\theta) \left(\log(q(\theta)) - \log(p(\theta | z))\right) d\theta.
\end{equation}
As $p(\theta | z)$ is typically intractable, VB maximises $\mathcal{L}(q)$ where
\begin{equation}
\label{ELBO}
\mathcal{L}(q, \lambda) = \int_{\theta} q(\theta) \left(\log(p(\theta, z)) - \log(q(\theta))\right) d\theta
\end{equation}
which is equivalent to minimising (\ref{KL-div}), where $\lambda$ is the parameter vector of $q(\theta)$.
This is done with stochastic gradient ascent, taking Monte Carlo estimates of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$ (Hoffman et al. 2013)
and repeating updates of the form
\begin{equation}
\label{gradientAscent}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} 
\end{equation}
which is guaranteed to converge to a local maximum (Robbins and Munro, 1951) if $\rho^{(m)}, m = 1, \dots, \infty$ satisfies
\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}
In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of Kingma and Ba (2015).

There are two Monte Carlo estimators of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the score estimator of Ranganath, Gerrish and Blei (2014), 
\begin{equation}
\label{scoreDeriv}
\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^L \frac{\delta \log(q(\theta_i))}{\delta \lambda} \left(\log(p(\theta_i, z)) - \log(q(\theta_i)) \right), \mbox{ where } \theta_i \sim q(\theta),
\end{equation}
and the reparameterised estimator of Kingma and Welling (2014),
\begin{equation}
\label{rpDeriv}
\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^M \frac{\delta f(\lambda, \epsilon_i)}{\delta \lambda} \frac{\delta \log(p(\theta, z))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_i)} + \frac{\delta J(\lambda, \epsilon_i)}{\delta \lambda}, \mbox{ where } \epsilon_i \sim r(\epsilon)
\end{equation}
where $r(\epsilon)$ is some distribution without any free parameters (eg. Uniform, Standard Normal), $f(\lambda, \epsilon)$ is some differentiable function that transforms $\epsilon$ to $\theta$, and $J(\lambda, \epsilon)$ is the Jacobian Matrix of this function. (\ref{rpDeriv}) typically has significantly lower variance than (\ref{scoreDeriv}), allowing $M$ to be set lower than $L$ which improves the speed of the algorithm; however the requirement for $f$ to exist restricts the class of approximating distributions $q(\theta)$ that can be used.
Notably, a differentiable $f(\lambda, \epsilon)$ does not exist when $q(\theta)$ is a mixture model and the score estimator (\ref{scoreDeriv}) must be used. To offset the increased variance and allow a reduction in $L$, following Gunawan, Tran, and Kohn (2017), Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence (Sobol 1967), then these are randomised using the scrambled net method (Matousek 1998), and transformed into a draw of $\theta$ using the inverse-CDF of $q(\theta)$. Using RQMC allows (\ref{scoreDeriv}) to converge to the true derivative at a rate $O(L)$ compared to $O(\sqrt{L})$ in standard Monte Carlo.

\subsection{Updating VB Process}
\item While on the road a self driving car may wish to make posterior inferences on the parameter set of cars around them. They may be interested in the degree of erraticness / variance of their actions.
\item The posterior for car $N+1$ is given by 

@@

\label{originalPost}
p(\theta_{N+1} z_{N+1} | z_{1:N}) \propto p(z_{N+1} | \theta_{N+1}) \int_{\beta} p(\theta_{N+1} | \beta) p (\beta | z_{1:N}) d\beta

@@

\item As N is large $p (\beta | z_{1:N})$ is fairly precise, so we can replace $\int_{\beta} p(\theta_{N+1} | \beta) p (\beta | z_{1:N}) d\beta$ with $ p(\theta_{N+1} | \hat{\beta})$ where $\hat{\beta}$ is the posterior mean of $p(\beta | z_{1:N})$. This approximation gives us a closed form for $ p(\theta_{N+1} | \hat{\beta})$.
\item Subsequently fitting $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})  \propto p(z_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \hat{\beta})$ is fairly easy with MCMC / VB etc.
\item When $S$ is small, the information in the prior will dominate the individual information.
\item However, for some $T > S$ the information in the car's individual movements should become more relevant and forecasts could be improved by updating the distribution to $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:T})$.
\item This updated posterior is given by

@@

\label{updatePost}
p(\theta_{N+1} | z_{N+1, 1:T}, z_{1:N}) \propto p(z_{N+1, S+1:T} | \theta_{N+1})p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})

@@

\item Both $\mathcal{L}(q)$ derivative estimators, (\ref{scoreDeriv}) and (\ref{rpDeriv}), need to evaluate \newline $\log(p(\theta_{N+1}, z_{N+1, S+1:T} | z_{1:N}, z_{N+1, 1:S}))$, which takes the form
\log(p(\theta_{N+1}, z_{N+1, S+1:T} | z_{1:N}, z_{N+1, 1:S})) &= \log(p(z_{N+1, S+1:T} | \theta_{N+1})) \nonumber \\
&+ \log(p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})) \label{updateEq}
\item To perform this update only using the data from $z_{N+1, S+1:T}$ we would need a closed form form expression for $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})$ to replace the role of the prior. Without this closed form we would need to re-fit the VB model using the original prior $p(\theta_{N+1} | \hat{\beta})$ and the whole observed sequence $z_{N+1, 1:T}$.
\item Due to the demands of driving, the car wants the update carried out as soon as possible, and ideally should be able to discard the previously used information $z_{N+1, 1:S}$. 
\item Defining $q(\theta_{N+1} | z_{N+1, 1:S})$ as the Variational Bayes approximation of $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})$, a new method is proposed to update $q(\theta_{N+1} | z_{N+1, 1:S})$ to $q(\theta_{N+1} | z_{N+1, 1:T})$ after observing $z_{N_1, S+1:T}$.
\item By replacing $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})$ in (\ref{updateEq}) with the previously fit $q(\theta_{N+1} | z_{N+1, 1:S})$ we get an approximation to (\ref{updateEq}):
\label{updateEqApprox}
\log(p(\theta_{N+1}, z_{N+1, 1:T} | z_{1:N})) \approx \log(p(z_{N+1, S+1:T} | \theta_{N+1})) + \log(q(\theta_{N+1} | z_{N+1, 1:S}))
\item Substituting (\ref{updateEqApprox}) into (\ref{scoreDeriv}) gives the score estimator for updating Variational Bayes, 
\label{scoreUpdate}
\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^L \frac{\delta \log(q(\theta_{N+1, i} | z_{N+1, 1:T}))}{\delta \lambda} &\left(\log(q(\theta_{N+1, i} | z_{N+1, 1:S}) - \log(q(\theta_{N+1} | z_{N+1, 1:T})) \right), \nonumber \\
& \mbox{ where } \theta_{N+1, i} \sim q(\theta_{N+1} | z_{N+1, 1:T}).
\item Similarly, substituting (\ref{updateEqApprox}) into (\ref{rpDeriv}) gives the reparameterised estimator for updating Variational Bayes,
\label{rpUpdate}
\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T} \approx \sum_{i = 1}^M \frac{\delta f(\lambda_T, \epsilon_i)}{\delta \lambda_T} \frac{\delta \log(q(\theta_{N+1} | z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon_i)} + \frac{\delta J(\lambda_T, \epsilon_i)}{\delta \lambda_T}, \mbox{ where } \epsilon_i \sim r(\epsilon)
where $\lambda_T$ denotes the parameter vector of $q(\theta_{N+1} | z_{N+1, 1:T})$.
\item This process can be repeated as neccessary, and is $O(1)$ instead of $O(T)$ assuming $S$ grows at the same rate as $T$.

\section{Forecast Evaluation}

\begin{itemize}
\item 4071 vehicles are available in the sample
\item 500 are used to estimate splines and straighten the road to relative coordinates as described in Section 2
\item 2000 are used to train any models used
\item 1571 are available for testing
\item Each second (10 observations) models are fit to the sample of $z_{N+1, 1:T}$ and forecasts of $z_{N+1, T+1:T+30}$ are made, corresponding to a three second forecast horizon
\item Forecasts are compared in terms of predictive log-score (where a predictive density is available) and Euclidean error.
\end{itemize}

\subsection{Heterogenous Time Series Models}

>>>>

\def\app#1#2{%
  \mathrel{%
    \setbox0=\hbox{$#1\sim$}%
    \setbox2=\hbox{%
      \rlap{\hbox{$#1\propto$}}%
      \lower1.1\ht0\box0%
    }%
    \raise0.25\ht2\box2%
  }%
}
\def\approxprop{\mathpalette\app\relax}

@@

\title{Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory}

%TODO:
%Flesh out section 1
%Add authors, institutions, an abstract and bibtex references 
%Switch out manual references for Bibtex when everything else is done.

\begin{document}
\maketitle

\section{Introduction}
\label{sec:intro}

@@

\item Self driving vehicles are becoming more common, eg Waymo, Tesla, GM.
\item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the self driving vehicle.
\item While self driving vehicles reduce collisions due to human error, safety can be further improved by better forecasts of the trajectory of nearby human controlled vehicles.
\item These models use deep neural networks (DNN) to extract the properties of surrounding vehicles from sensors (eg. Cameras and LiDAR), such as their position, velocity, and steering angle, see \citet{Woo2016a} and \citet{Tian2017}.
\item The self driving vehicle forecasts point estimates of the future trajectory of these surrounding vehicles based on the DNN results.
\item These can assume the surrounding vehicles will maintain their current angle and velocity \citet{Gindele2010}
\citet{Waymo2017}, \citet{Bautista2017}, but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citet{Woo2016b}, \citet{Ding2013}, \citet{Woo2017}, \citet{Geng2017}.

@@

\item An important part of this modelling is to capture the uncertainty of car trajectories, as the self driving vehicle must react to the possibility of collision even when that possibility is relatively small.

@@

\item This model implies that any new vehicle observed by the self driving vehicle has its own parameter vector, related to previously observed vehicles through the shared prior.
\item The self driving vehicle must be able to estimate the parameter vector of newly observed cars, and update this parameter vector as it observes additional data.

@@

\item Fit a hierarchical model with a large amount of vehicles to obtain the posterior distribution of each vehicles individual parameter vector and the shared distribution. This can occur before the self driving vehicle is on the road.

@@

\item Self driving vehicles take in up to 1GB of data a second (several websites report this as quotes from different people eg. Tech CEO, but I don't know if that's good enough for an academic citaiton), so discarding old data is useful.

@@

\section{Data processing}
\label{sec:dataProcessing}
Data is provided by the Next Generation Simulation (NGSIM) project, which recorded vehicles travelling along a 2235 feet section of the US 101 Freeway in Los Angeles, California from 7:50 am to 8:35 am on June 15th, 2005. Data was collected by seven static cameras and processed by Cambridge Systematics Inc. The result is coordinates relative to the start of the road section for 6101 vehicles collected every 100 milliseconds. The average number of observations per vehicle is 672, corresponding to a 67.2 seconds of data.
Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which travel along the five major lanes or the enter from the entry/exit lane to the right, with one vehicle changing from Lane 2 to Lane 1. Vehicles that entered or exited midway through the freeway section are excluded in this paper. There is a curvature to the road, with bends occuring between 500 and 1000 feet and 1800 and 2000 feet.

\begin{figure}
\centering
\includegraphics[width=0.65\textwidth]{carPath}
\caption{The path of ten vehicles in the dataset. The freeway is split into five lanes, and an additional entry/exit lane to the far right, through which two vehicles from this sample have entered. There is curvature to the road slightly distored by the aspect ratio, with bends occuring around 500 and 1800 feet.}
\label{fig:rawData}
\end{figure}

Modern vehicles are capable of tracking lane marking, and thus can compensate for the curvature of the road as they calculate the location of surrounding vehicles in coordinates relative to their own position (Thuy and Leon 2010), rather than relative to a static point as provided by NGSIM. To compensate for this, and so that the data is in a coordinate system similar to that of a self driving vehicle tracking surrounding cars, the coordinate system provided in the NGSIM dataset is transformed into a relative system where the road curvature is removed. This is facilitated by measuring the location of each vehicle relative to the estimate of the midpoint of each lane. \citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to build a model of the lane midpoints for the entire road section.

The midpoints for each of the five lanes are estimated separately, using 100 randomly sampled vehicles each that did not change their lane over the observed period. For each sampled vehicle $i$ and time since entering the road $t$, the total distance travelled is calcualted as 

@@

\label{distance}
d_{i, t} = \sqrt{(x_{i, t}) - x_{i, 1})^2 + (y_{i, t} - y_{i, 1})^2}

@@

From this distance metric, the average $x$ and $y$ coordinates at each given lane and for a given distance are each estimated by a pair of smoothing splines calculated with the R `stats' package. The smoothing splines are functions of the distance travelled and estimate the lane midpoints as  $\hat{x}_{i, t} = f(d_{i, t})$ and $\hat{y}_{i, t} = g(d_{i, t})$. Each remaining vehicle in the dataset uses the lane midpoint estimates $\{\hat{x}^*_{i, t}, \hat{y}^*_{i, t}\}$ fit from the spline model assocaiated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the devitation from the lane midpoint, by
x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right) - \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
y^*_{i, t} &= d_{i, t}. \label{yRel}
The vehicles used to estimate the five spline models are excluded for any other purpose.
\begin{figure}
\centering
\includegraphics[width = 0.5\textwidth]{relCoord}
\caption{An example of the equation in (\ref{xRel}). The estimated trajectory of the lane midpoint is given by the solid curve, while a vehicle $i$ at a given point in time $t$ is given by the blue dot at $\{x_{i, t}, y_{i, t}\}$. This vehicle has travelled a distance equivalent to the red dot at $\{\hat{x}_{i, t}, \hat{y}_{i, t}\}$. The gradient of the midpoint at this point is given by the dashed line, which intercepts the x-axis at an angle of $\lambda = \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)$. The dotted line from the blue dot to the red intercepts the x-axis at an angle of $\psi = \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)$. The sign of the angle $\lambda - \psi$ determines whether the vehicle is to the left or right side of the lane, and thus has a negative or positive relative coordinate $x^*_{i, t}$. The absolute value of $x^*_{i, t}$ is the distance between the two dots.}
\label{fig:relCoord}
\end{figure}
Once relative coordinates are extracted, the relative position of any vehicle $i$ at any time $t$ is determined by its initial position $\{x^*_{i, 1}, y^*_{i, 1}\}$, as well as the history of that driver's inputs: velocity $\{v_{i, s} | s = 2, \dots, t\}$ and steering angle, $\{\delta_{i, s} | s = 2, \dots, t\}$ by the trigonometric relationship in Figure \ref{fig:motion}:
x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq} \\
y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq},
where $\delta_{i, t} = \pi/2$ denotes that the car has no lateral relative movement and $x^*_{i, t} = x^*_{i, t-1}$.
From this relationship, the driver's inputs are be extracted by
\delta_{i, t} &= \tan^{-1}\left(\frac{(y^*_{i, t} - y^*_{i, t-1})}{(x^*_{i, t} - x^*_{i, t-1})} \right) \label{dEq} \\
v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{vEq}.
It is then useful to extract the acceleration, $a_{i, t}$ by the simple relation

@@

\label{aEq}
a_{i, t} = v_{i, t} - v_{i, t-1}. 

@@

Data for vehicle $i$ up to time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $z_{1:N} = \{z_{i, 1:T_i} | i = 1, \dots, N\}$.
\begin{figure}
\centering
\includegraphics[width = 0.4\textwidth]{motion}
\caption{A vehicle at $x_{t-1}, y_{t-1}$ at time $t-1$ will be at $x_t, y_t$ at time $t$ if its angle over this period is $\delta_t$ and velocity is $v_t$.}
\label{fig:motion}
\end{figure}

\section{Trajectory Forecasts}
\label{sec:trajForecasts}

A crucial step in the safe operation of self driving vehicles is to forecast the future trajectory of surrounding vehicles so that collisions can be avoided. To produce these forecasts the data is split to place vehicles into two sets: a training set used to infer the posterior distribution of a parameter set before the self driving vehicle is on the road; and a test set used for forecasting the positions of vehicles observed while driving. The parameters associated with the vehicles in the test set may be inferred, but this inference is time constrained. If inference takes $M$ seconds, a forecast after observing a vehicle for $T$ seconds can only include data up to time $T - M$ to infer the posterior distribution of any parameters.

The remainder of Section \ref{sec:trajForecasts} will introduce the models used, while Sections \ref{sec:Inference} and \ref{sec:UVB} will build an inferential framework to facilitate inference of any additional parameters in constant time as $T$ increases to apply to online trajectory forecasting.

\subsection{Homogenous Time Series Model}
\label{subsec:homogenous}

The framework provided by (\ref{xEq}) and (\ref{yEq}) allows forecasts of $\{x^*_{t}, y^*_{t} | t = T + 1, \dots, T+H\}$ to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_T, y^*_T, v_T\}$, and forecasts of the future values of 
$\{a_{t}, \delta_{t} | t = T + 1, \dots, T+H\}$.

As data is observed every 100 milliseconds, large changes in $a_{i, t}$ and $\delta_{i, t}$ can occur over multiple time periods. This can be observed in Figure \ref{fig:dynamics}, where the left panel plots the mean value of
\begin{equation}
A_{i, t} = \frac{a_{i, \tau_i + t}}{a_{i, \tau_i}}
\label{amax}
\end{equation}
where $\tau_i = \arg \underset{t}{\max}|a_{i, t}|$. On average, a vehicle will increase in acceleration for the 400 milliseconds before reaching maximum acceleration. The right panel plots the mean value of
\begin{equation}
D_{i, t} = \frac{\delta_{i, \omega_i + t} - \pi/2}{\delta_{i, \omega_i} - \pi/2}
\label{dmax}
\end{equation}
where $\omega_i = \arg \underset{t}{\max}|\delta_{i, t} - \pi/2|$, which similarly demonstrates that large absolute deviations in steering angle from $pi/2$ are associated with smaller deviations in the same direction in the the previous 300 milliseconds.

\begin{figure}[h]
\centering
\includegraphics[width = 0.95\textwidth]{dynamics}
\caption{Left: Mean values of acceleration as a percent of a vehicles maximum absolute acceleration for the 1000 milliseconds before, and 500 milliseconds after, a vehicle reaches its maximum absolute acceleration. Right: The analogous plot for the steering angle, with $\delta_t - \pi/2$ as a percentage of a vehicles maximum absolute value of $\delta_t - \pi/2$. The few observations before a variable reaches its maximum often have increasing values for the same variable.}
\label{fig:dynamics}
\end{figure}

These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and thus auto-regressive processes of orders $p$ and $q$ for $a_t$ and $\delta_t$ are employed to model this, with
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{j} a_{i, t-j} + \sigma_{\epsilon} \epsilon_{i, t} \label{aAR} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta} \eta_{i, t} \label{dAR}
\end{align}
for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution.

The parameter vector is transformed to $\mathbb{R}^{p + q + 2}$ and collected as 
\begin{equation*}
\label{thetaVec}
\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\},
\end{equation*}
where
\begin{equation}
\label{indPrior}
p(\theta) \sim \mathcal{N}\left(\mu, \Sigma \right)
\end{equation}
where $\mu$ is a vector with an entry of $0$ for each $\phi$ and $\gamma$, and an entry of $-5$ for each log-variance, as the scale of changes in the data is very small. $\Sigma = 10 \mathbb{I}$, where $\mathbb{I}$ is the identity matrix. This is referred to as the homogenous model throughout the paper.

\subsection{Heterogenous Time Series Models}
\label{subsec:heterogenous}
